Correlation tests

### 2025-04-04 analysis/manuscript.qmd
Rough intro outline

This study aims to demonstrate the suitability of open scholarly data sources for assessing the impact of transformative agreements on hybrid open access.
To achieve this, a replication study was conducted by comparing results from hoaddata, an openly available and continuously updated dataset on hybrid open access uptake based on Crossref, OpenAlex, and the cOAlition S Journal Checker Tool, with the established bibliometric databases Web of Science and Scopus.

This study focuses on the coverage of hybrid journal portfolios included in transformative agreements between 2019 and 2023.
Special attention is given to potential differences in open access uptake by country when comparing first-author affiliation data to corresponding authorships.
This is crucial because the lack of publicly available invoicing data corresponding to authorships plays an essential role in determining whether an open-access article is supported through transformative agreements.
Data on corresponding authorships have been available on the Web of Science and Scopus for much longer than in open databases such as OpenAlex, where this information is still being roled out at the time of writing.
Because of this weakness, open approaches such as hoaddata and related research use first-authorship data instead.

By conducting a large-scale comparative analysis, this study aims to

1. Determine the strengths and weaknesses of using open data sources in monitoring the impact of transformative agreements on hybrid open access publishing.
2. Assess the coverage and accuracy of open data sources compared with established bibliometric databases.
3. Evaluate the reliability of first author affiliation data as a proxy for corresponding authorship in the context of open access uptake analysis.

# Background -- Evidence base to measure the effects of transformative agreements

## Anforderungen an das Monitoring

- esac guidelines 
- gemeinsamkeiten und unterschiede zu apc (listenpreise, tatsächliche zahlungen, zentrales invoicing, rabatte, waivers) 
- insitutionen covern cas, jedoch kann es zu unterschiedlichen verrechnugnsformen führen (antielig mit förderer, splitting innerhaklbd er einrichtung)

## Bibliometrische Evidenzen

- allgmeeiner uptake 
- wachstum apcs 
- wachstum verträge (konsortien, forschung) 
- konsequenzen

@@

This study of over 13,000 hybrid journals included in the Transformative Agreement Data underlying the cOAlition S Journal Checker Tool shows a rise in open access due to transformative agreements between 2019 and 2023, although most articles remained paywalled.
While transformative agreements accounted for the majority of open access, many articles continue to become open through the payment of individual publication fees.
Hybrid open access and transformative agreements remain concentrated among a small group of large commercial publishers, 
with high-income European countries—alongside South Africa—showing high adoption rates.
In contrast, the three most productive countries, China, the United States, and India, were ranked much behind other countries in the adoption of transformative agreements.
Open questions remain whether this uneven distribution reflects temporary implementation gaps, inherent inequities in the transformative agreement model, or countries voluntarily refraining from such agreements.

These trends were consistent across the investigated open data source hoaddata, derived from Crossref and OpenAlex, and the established proprietary bibliometric databases Scopus and Web of Science. 
Notably, the results show strong correlations by country affiliation, aligning with previous studies [@Akbaritabar_2024; @alperin2024analysissuitabilityopenalexbibliometric], which supports the use of open research information for large-scale analyses of hybrid open access. 
However, the observed differences in coverage of journals, articles, and metadata availability warrant discussion, affecting not only open data sources but also proprietary databases when used in isolation.

To begin with, the coverage analysis reveals that hybrid journals are well indexed in all three data sources, particularly in terms of article coverage.
Differences can be found for journals targeting practitioners or local non-English language communities, with many such titles indexed exclusively in Crossref and Scopus.
Using the open index Crossref as a bibliometric database in hoaddata demonstrated particular strength in identifying newly established hybrid journals, a notable finding given that transformative agreements primarily target existing subscription-based journals.

The situation of hybrid journal publishing is therefore different from that of fully open access journals.
Comparing the coverage of OpenAlex, Scopus and Web of Science, @Simard_2024 indicate that only half of the fully open access journals listed in the DOAJ are also indexed in Scopus and Web of Science.
Notably, journals that charge no publication fees ("diamond journals") are absent from the selective Web of Science, which reinforces existing disparities in the indexing of underrepresented research communities and regions in selective bibliometric databases (ref).

@@

One often reported limitation of studying open access with less selective databases is the lack of corresponding authorship information [@Fraser_2023; @Haucap_2021; @Shu_2023]. 
However, indicators based on first authors, which have often been used as a proxy for determining open access funding, and corresponding authors show a high level of correlation, reflecting disciplinary norms in scholarly publishing with regard to contributorship and author roles and positions [@Larivi_re_2016]. 
First authors are often considered to have carried out the main research underlying a paper, while the corresponding author is often recognised as having supervised the research [@Mattsson_2010; @Fox_2018]. 
Unsurprisingly, measures based on first or corresponding authorships are strongly correlated, suggesting that these authors share the same country affiliation. Moreover, in most cases, the first author is identical the corresponding author [@Chinchilla_Rodr_guez_2024].
Thus, first authorship can reliably be used in large-scale studies to assess the impact of transformative agreements.

Despite this correlation, the recent slight increase in open access articles by corresponding authors over first authors across Springer Nature hybrid journals prompts questions about how institutional open access sponsorship practices influence author roles and assignments within co-author teams, especially as funding opportunities vary [@Gumpenberger_2018]. Previous research has highlighted how institutionalized bibliometric practices can affect the valuation of authorship positions [@Helgesson_2020], underscoring the need to be cautious of monetary authorship influences [@Maddi_2024].

Affiliation data is another important data element for estimating open access enabled by transformative agreements. 
Although OpenAlex's affiliation coverage is less comprehensive, which likely reduced the number of articles confidently attributed to transformative agreements in this study, it still shows high correlations with Scopus and Web of Science at the country level.
However, OpenAlex's native ROR ID integration offers a distinct advantage, enabling more reliable identification of agreement-enabled articles compared to Scopus and Web of Science, which require reconciliation with proprietary organization identifiers. 
Future studies based on the Web of Science will benefit of the recent integration of ROR-IDs, announced by the end of 2024.
<!-- Dimension is another scholarly data source that uses open standards to identify journals, which enables analysing transformative agreements [@Fraser_2023].-->

Comparing scholarly databases reveals discrepancies in open access evidence. Not all publishers deposit Creative Commons (CC) licenses with Crossref, but comparing it with Unpaywall (integrated into Scopus and Web of Science) exposes these gaps, as these discovery services also analyze publishers' websites.

Comparing scholarly databases helps to identify  discrepancies in open access evidence. 
Crucially, not all publishers deposit CC licences with Crossref, a limitation that becomes apparent when contrasting Crossref with Unpaywall’s data in Scopus and Web of Science. 
While Unpaywall can detect such gaps by parsing publishers' websites for open access licences, it nevertheless missed a substantial number of CC-licensed open access articles from Wiley journals indexed in Crossref, likely due to parsing errors on the publisher's website. This resulted in fewer open access articles being recorded in Scopus and Web of Science.
Further inconsistencies could be observed between Scopus and Web of Science, despite both relying on Unpaywall: Web of Science erroneously labels Elsevier’s delayed OA as hybrid, whereas Scopus correctly categorises it.
ONotably, Scopus and Crossref showed greater alignment than observed in a related comparison between Scopus and OpenAlex [@alperin2024analysissuitabilityopenalexbibliometric]. This likely reflects the use of a dedicated list to identify hybrid open access journals, rather than relying solely on article-level tags.
Given these variations, which may compromise comparability, research and monitoring exercises should avoid reliance on a single source. Instead, the selection of open access evidence should be cross-verified using multiple sources [@Huang_2020]. Incorporating expert-curated journal lists can also help prevent misclassifications based on business models [@Visser_2021].

It is important to note that articles from institutions involved in transformative agreements are estimates due to a lack of access to invoice data, which is not usually shared between library consortia and publishers. 
Furthermore, it is important to acknowledge that the sources of transformative agreements and the underlying ESAC registry are a voluntary effort, crowd-sourced from various consortia. 
However, recent validation with Dutch Research Information demonstrates the reliability of such an open approach for the assessment of articles under transformative agreements. 
With enhanced and streamlined metadata compliance, attributable to evolving standards and initiatives to support negotiations with publishers, particularly through the communities of the ESAC initiative and Barcelona Declaration supporters, the situation is likely to improve. In conclusion, this cross-database analysis has highlighted the potential and challenges of establishing open scholarly data as a suitable evidence base for analysing the transition from subscription-based publishing to open access.

>>>>

Transformative agreements have emerged as a much-discussed licensing model for the transition from subscription-based publishing to full open access in the last ten years. 
While these agreements can vary substantially, increasing coordination exists in standardising the negotiation, administration, and evaluation of these agreements. 
Simultaneously, efforts to improve transparency through open data about contracts and payments are growing. 
However, this information remains only partially available through bibliometric databases, limiting large-scale studies about the impact of transformative agreements on the transition of hybrid journals to full open access through increased open access uptake.
This study compares findings from hoaddata, an openly available,  continuously updated dataset on hybrid open access, built on open data from Crossref, OpenAlex and the cOAlition S Journal Checler Tool, with the established bibliometric databases Web of Science and Scopus. 
The purpose of this comparison is not only to evaluate the suitability of open metadata for measuring hybrid open access but also to inform bibliometric research and practice by highlighting the strengths and weaknesses of each data source.

Since their introduction, transformative agreements have sparked considerable academic debate. 
Whilst standardisation activities, particularly through the OA2020 and ESAC (Efficiency and Standards for Article Charges) initiatives, have led to improved workflows between library consortia and publishers, as well as increased open access publishing, the initial focus on a few large publishers by national consortia from high-income countries has drawn criticism. 
Critics have raised concerns about market concentration, the lack of journal flipping to full open access models, reduced competition in the publishing market, and wider gaps between well-resourced and under-resourced institutions and researchers.

For instance, studying over 700 agreements, Jahn (2025) observed a strong growth in open access publishing to 15% in 2022 due to the varying implementation of transformative agreements across countries, predominantly driven by three commercial publishers: Elsevier, Springer Nature, and Wiley.
This aligns with the rapid proliferation of transformative agreements in recent years.
Previously, only a minimal proportion of articles were made open access in hybrid journals (Piwowar, 2018; Laakso, 2016).

Evaluations conducted by library consortia confirm these trends but draw different conclusions.
For instance, the British Joint Information Systems Committee's evaluation (Brayman, Devenney et al., 2024) noted transformative agreements' significant impact on national open access growth but limited effect on global open access transition, suggesting a need to reevaluate their strategy.
Similar policy reconsiderations emerged in Norway and Sweden (Holden, Skoie et al., 2023; Widding, 2024).
Meanwhile, cOAlition S recommended ending financial support for these agreements in 2024, but still considers hybrid open access compliant with funder open access policies.
Conversely, Germany's DEAL consortium extended agreements with major publishers through 2028, while Colombia established Latin America's first transformative agreements (Muñoz-Vélez, Pallares et al., 2024). A recent Max Planck Digital Library study demonstrates how cost savings can be achieved through this model and how it can be further developed to achieve full open access while controlling expenditure.

Open databases like OpenAlex, which build upon Crossref, particularly benefit from these developments.
These resources have gained considerable attention in research evaluation due to their inclusivity, openness, and comprehensive coverage of publication metadata.
Recent studies [REFERENCE NEEDED] have demonstrated that open bibliometric databases increasingly match or exceed the coverage of traditional commercial databases in many disciplines, whilst providing improved access to standardised and interlinked open research information.

Against this background, this study evaluates the suitability of open metadata to measure the impact of transformative agreements on hybrid open access.
Only a few studies have attempted this analysis, primarily due to data limitations and the need to combine multiple sources.
Transformative agreements present unique measurement challenges because of their complex workflow requirements, which are not fully reflected in bibliometric databases.
These challenges include varying journal portfolios across agreements, invoicing procedures specific to submitting authors, and article caps that limit open access publishing within agreements.

The present study addresses these gaps by replicating and applying consistent methodologies across Web of Science, Scopus, and open data sources.
This comparative approach enables the research to overcome the limitations of any single source whilst systematically assessing the strengths and weaknesses of different bibliometric data sources for monitoring transformative agreement outcomes.
By triangulating findings across these platforms, the study aims to provide a more comprehensive understanding of how transformative agreements influence hybrid open access publishing patterns within the ongoing scholarly publishing transition.

@@

This study of over 13,000 hybrid journals shows a rise in open access due to transformative agreements between 2019 and 2023, although most articles remained paywalled.
While transformative agreements accounted for the majority of open access, many articles continue to become open through the payment of individual publication fees.
Hybrid open access and transformative agreements remain concentrated among a small group of large commercial publishers, 
with high-income European countries—alongside South Africa—showing high adoption rates.
In contrast, the three most productive countries, China, the United States, and India, were substantially ranked lower in transformative agreement adoption.
Open questions remain whether this uneven distribution reflects temporary implementation gaps, inherent inequities in the transformative agreement model, or countries deliberately refraining from such agreements.

The findings were consistent across the investigated open data source hoaddata, derived from Crossref and OpenAlex, and the established proprietary bibliometric databases Scopus and Web of Science. 
Notably, the results show strong correlations by country affiliation, aligning with previous studies [@Akbaritabar_2024; @alperin2024analysissuitabilityopenalexbibliometric] and rankings [@eck_2024_6bx73], which supports using open metadata for large-scale analyses of hybrid open access. 
However, the observed differences in journal coverage and metadata availability warrant discussion, affecting not only open data sources but also proprietary databases when used in isolation.

The coverage analysis reveals that hybrid journals are well indexed in all three data sources, particularly in terms of article coverage.
Differences emerge for journals targeting practitioners or local non-English language communities, with many such titles indexed exclusively in Crossref and Scopus.
Using the open index Crossref as a bibliometric database in hoaddata demonstrated particular strength in identifying newly established hybrid journals, a notable finding given that transformative agreements primarily target existing subscription-based journals.

The landscape of hybrid journal publishing differs markedly from that of fully open access journals.
Comparing the coverage of OpenAlex, Scopus and Web of Science, @Simard_2024 indicate that only half of the fully open access journals listed in the DOAJ are also indexed in Scopus and Web of Science.
Notably, journals that charge no publication fees ("diamond journals") are absent from the selective Web of Science, which reinforces existing disparities in the indexing of underrepresented research communities and regions in selective bibliometric databases (ref).

@@

A frequently reported limitation of studying open access with less selective databases is the lack of corresponding authorship information [@Fraser_2023; @Haucap_2021; @Shu_2023]. 
However, this analysis demonstrates that indicators based on first authors, which have often been used as a proxy for determining open access funding, and corresponding authors show a high level of correlation, reflecting disciplinary norms in scholarly publishing with regard to contributorship and author roles and positions [@Larivi_re_2016]. 
First authors typically conduct the main research underlying a paper, while the corresponding author often supervises the research [@Mattsson_2010; @Fox_2018]. 
Unsurprisingly, measures based on first or corresponding authorships are strongly correlated, suggesting that these authors share the same country affiliation. Moreover, in most cases, the first author is identical to the corresponding author [@Chinchilla_Rodr_guez_2024].
Thus, first authorship can reliably be used in large-scale studies to assess the impact of transformative agreements.

Despite this correlation, the study observed a recent slight increase in open access articles by corresponding authors over first authors across Springer Nature hybrid journals prompts questions about how institutional open access sponsorship practices influence author roles and assignments within co-author teams, especially as funding opportunities vary [@Gumpenberger_2018]. 
Previous research has highlighted how institutionalized bibliometric practices can affect the valuation of authorship positions [@Helgesson_2020], suggesting the need to monitor potential influences on authorships [@Maddi_2024].

Another critical data element in the study is affiliation data, which is essential for estimating open access enabled by transformative agreements. 
Although OpenAlex's affiliation coverage is less comprehensive, which likely reduced the number of articles confidently attributed to transformative agreements in this study, it still shows high correlations with Scopus and Web of Science at the country level.
However, OpenAlex's native ROR ID integration offers a distinct advantage, enabling more reliable identification of agreement-enabled articles compared to Scopus and Web of Science, which require reconciliation with proprietary organization identifiers. 
Future studies based on the Web of Science will benefit from the recent integration of ROR-IDs, announced by the end of 2024.
<!-- Dimension is another scholarly data source that uses open standards to identify journals, which enables analysing transformative agreements [@Fraser_2023].-->

The database comparison revealed important discrepancies in open access evidence. 
Crucially, not all publishers deposit CC licences with Crossref, a limitation that becomes apparent when contrasting Crossref with Unpaywall’s data in Scopus and Web of Science. 
While Unpaywall can detect such gaps by parsing publishers' websites for open access licences, it nevertheless missed a substantial number of CC-licensed open access articles from Wiley journals indexed in Crossref, likely due to parsing errors on the publisher's website. This resulted in fewer open access articles being recorded in Scopus and Web of Science.
Further inconsistencies emerged between Scopus and Web of Science, despite both relying on Unpaywall: Web of Science erroneously labels Elsevier’s delayed open access as hybrid, whereas Scopus correctly categorises it.
Notably, Scopus and Crossref showed greater alignment than observed in a related comparison between Scopus and OpenAlex [@alperin2024analysissuitabilityopenalexbibliometric]. This likely reflects the use of a curated list to identify hybrid open access journals, rather than relying solely on article-level tags.
Given these variations, which may compromise comparability, research and monitoring exercises should avoid reliance on a single source. Instead, the selection of open access evidence should be cross-verified using multiple sources [@Huang_2020]. Incorporating expert-curated journal lists can also help prevent misclassifications based on business models [@Visser_2021].

It is important to note that the study's estimates of articles from institutions involved in transformative agreements are approximations due to a lack of access to invoice data, which is not usually shared by library consortia and publishers. 
Furthermore, the data sources about transformative agreements including the cOAlition S Journal Checker Tool and the ESAC registry are a voluntary effort, crowd-sourced from various consortia. 
However, recent validation with Dutch Research Information demonstrates the reliability of such an open approach for the assessment of articles under transformative agreements [@de_Jonge_2025]. 
With enhanced and streamlined metadata compliance, attributable to evolving standards and initiatives to support negotiations with publishers, particularly through the communities of the ESAC initiative and Barcelona Declaration supporters, the situation is likely to improve. In conclusion, this study has shown that open scholarly data are suitable for analysing the transition from subscription-based publishing to full open access.
